Mapping Mixtures of Genotype-by-Weather Interactions in Switchgrass (Panicum virgatum)

### 2024-01-08 manuscript/revision2/revision2.qmd
manuscript: revision 2 text work Abstract through Discussion

title: Switchgrass Phenology and Genotype-by-Environment Mapping

@@

  The timing of vegetative growth and flowering in plants ("phenological timings") depend on both genetic variation and environmental cues; thus, phenological traits can have important genotype-by-environment (GxE) interactions. In particular, understanding the mixture and prevalence of genotype-by-weather (GxWeather) interactions primed by weather prior to the phenological event should aid prediction and manipulation of phenological timings. Here, we map genotype-by-weather effects on phenological timings in two highly divergent switchgrass (*Panicum virgatum*) populations using repeated plantings of these populations at sites spanning the central United States. We distinguish GxWeather patterns covarying with interpretable weather-based cues from agnostic, site-based patterns. Most GxWeather effects belong to the latter category and do not covary with weather-based cues. However, 65% of Gulf population effects on the start of vegetative growth covary with daylength 14 days prior to green-up date. 33% of Gulf population effects on flowering date covary with cumulative rainfall in the seven days prior to flowering, while 22% of Midwest population effects on flowering date covaried with day length change two days prior to flowering. An independent pseudo-F2 cross of Gulf and Midwest individuals mapped 23 additive QTLs for flowering at the same common gardens, all with significant mash associations and eleven with enrichment of highly significant mash associations. We demonstrate that we can identify QTL with GxWeather and assign them to specific weather-based cues or other patterns. Breeding for particular alleles at these loci could change flowering responsiveness to photoperiod cues in switchgrass. More broadly, this approach could be used to identify genetic marker-environment interactions in any species with related populations phenotyped in multiple environments.

@@

  The timing of plant seasonal development (phenology) has major impacts on fitness because of the steep price of plant-environment mismatches. We infer the mixture of ways genetic effects on phenological traits depend on plant environment, particularly how effects covary with weather prior to the phenological event (GxWeather). Most effects do not have GxWeather, but a minority do. GxWeather is population-specific. The majority of effects on the start of vegetative growth in the Gulf population have GxWeather: these effects differ in sign across environments and covary with a photoperiod cue 14 days prior. A minority of effects on flowering date in the Gulf and Midwest populations had GxWeather, covarying with a cumulative rainfall and a photoperiod cue, respectively.

@@

\firstpage[10]{2}

The timing of plant vegetative and reproductive development are major components of plant fitness affected by multiple external environmental cues (e.g. degree of winter chilling, day length, temperature, and water availability) that signal existing or upcoming growing conditions [@bauerle_photoperiodic_2012; @andres2012genetic; @korner2010phenology]. Genetic responses to environmental cues determine the speed, timing, and energy apportioned to vegetative and reproductive growth and shape both the individual's lifespan and its lifetime production of viable seed. Day length (or photoperiod) is one of the most predictable environmental cues, and genetic sensitivity to photoperiod protects plants from potentially fatal consequences of phenological responses to temperature cues at the "wrong" time of year. However, the usefulness of specific environmental cues depends on both features of the environment and the species' adaptive strategies @botero2015evolutionary. Species with wide natural distributions can have multiple distinct environmentally-cued phenological responses: for example, populations of sunflower (Helianthus annuus) exhibit day-neutral, facultative short day, and facultative long-day flowering responses, which vary with their environments [@blackman2013interacting; @henry2014transitions]. Distinct genetic responses in different environments are known as genotype by environment interactions, or GxE.

Flowering time, in particular, is a common subject of GxE research [@blackman2013interacting; @henry2014transitions; @aagren2017adaptive; @brachi2010linkage; @dittmar2014flowering; @li2018genomic; @romero2017study], a key output of selection driving adaptation to local environments [@andres2012genetic; @wadgymar2017identifying; @blumel2015flowering], and a selection target for crop improvement to adapt crops to local or future environments @jung2009flowering. Changing flowering responsiveness to photoperiod cues has allowed geographic range expansion and increased yields in a number of cereal species [@turner2005pseudo; @faure2012mutation; @hung2012zmcct; @zakhrabekova2012induced; @yang2013oself3] and other crops [@pin2012multifaceted; @weller2019parallel]. Recent statistical advances in studying phenological GxE have involved determining critical environmental indices before the phenological event occurs, such as photothermal time within a critical growth window @li2018genomic. However, most studies of flowering GxE focus on finding a single, best fitting form of genotype-environment covariance, despite the key expectation that different genetic subpopulations, and even different genomic regions, have likely evolved distinct patterns of GxE. Additionally, despite theoretical predictions that local adaptation should involve antagonistic pleiotropy, or sign-changing GxE, at the level of individual loci [@levene1953genetic; @felsenstein1976theoretical; @kawecki2004conceptual; @hedrick1986genetic], previous work has found limited evidence of antagonistic pleiotropy [@des2013genotype; @wadgymar2017identifying], but has been limited by a known statistical bias that reduced detection of genetic effects that differ in sign [@des2013genotype; @anderson2013genetic; @anderson2011evolutionary]. Thus, despite substantial interest in the frequencies of various forms of GxE, the prevalence of antagonistic pleiotropy relative to other forms of GxE remains unknown.

Switchgrass (*Panicum virgatum*) is considered a short-day plant with reproductive development strongly linked to day of the year @mitchell1997predicting. However, as part of its wide environmental adaptation across the eastern half of North America, its photoperiodicity has been predicted to differ by plant latitude of origin [@parrish2005biology; @casler2004latitudinal]. We previously found divergent Midwest and Gulf genetic subpopulations of switchgrass which segregate for distinct sets of environmental adaptations, based on the analysis of two measures of fitness (i.e.,biomass and survival) among 732 genotypes @lovell2021genomic. The Midwest genetic subpopulation is primarily composed of individuals from the well-studied upland switchgrass ecotype [@porter1966analysis; @milano2016genetic], while the Gulf subpopulation has individuals from the lowland ecotype and the phenotypically intermediate coastal ecotype @lovell2021genomic.

Here, we test how the Midwest and Gulf subpopulations differ in their phenological adaptations across environments, and hence their phenological GxE. We phenotype a diversity panel of hundreds of switchgrass genotypes from the Midwest and Gulf subpopulations for the timing of vegetative development ("green-up") and reproductive development (flowering) at eight common garden locations spanning 17 degrees of latitude. These gardens cover the majority of the latitudinal and climatic range of switchgrass and capture the most comprehensive picture to date of the environmental variation this species encounters. After single-site estimation of SNP effects for these phenotypes, we use multivariate adaptive shrinkage (mash) to jointly estimate SNP effects in each subpopulation and phenotype at all eight sites @urbut2019flexible. Mash allows us to identify and specify multiple ways genetic effects linked to SNPs may covary with the environment, and does not have a statistical bias in detecting SNP effects with the same or opposite signs @10.1093/biostatistics/kxw041. To confirm our genetic mapping of GxE, we compare to mapping results from an outbred pseudo-F2 cross grown at the same sites. Taken together, our results allow us to describe the environmental cues and genetic variation affecting phenology in two divergent natural populations of switchgrass.

@@

In our diversity panel of tetraploid switchgrass @lovell2021genomic, genotypes from the Gulf and Midwest genetic subpopulations had distinct phenological timings and distinct patterns of phenological correlations across our eight common garden sites (@fig-map). At the three Texas common gardens (hereafter 'Texas' gardens), located within the natural range of the Gulf subpopulation, Gulf green-up occurred before Midwestern green-up, and Gulf flowering occurred after Midwestern flowering (@fig-map A). At the four northernmost common gardens (hereafter 'North' gardens), located within the natural range of the Midwest subpopulation, both Gulf green-up and flowering occurred after Midwest green-up and flowering. At the Oklahoma common garden, located near the natural range limits of both the Gulf and the Midwest subpopulations, Gulf and Midwest green-up occurred over the same time period. These patterns led to strong negative phenotypic correlations for green-up between the North and Texas gardens, particularly in the Gulf and Both subpopulations, and contributed to positive phenotypic correlations for flowering time of larger magnitude at more northern gardens (@fig-map B).

@@

We next explored how genetic variation in phenology covaried with environmental cues by using mash to jointly discover trait-associated SNPs across all common gardens for the Midwest, Gulf, and 'Both' subpopulations (@fig-map D, SI Appendix, Datasets 1-6). We first estimated SNP effects using GWAS on site-specific BLUPs for each phenotype. We then used mash on two subsets of these SNP effects, first to determine the covariance structures that significantly improved the mash model log likelihood when included, then to re-estimate the set of SNP effects with the largest effect estimates in univariate GWAS (hereafter "strong" effects) using the set of significant covariance matrices.

Three types of covariance matrices were tested for inclusion: "canonical" covariance matrices, with simple patterns of effects; "data-driven" matrices derived from common patterns of SNP effects in the data, and "hypothesis-based" covariance matrices that captured the covariance of weather-based phenological cues for cloned genotypes grown at multiple common gardens (Table 1, SI Appendix, Section S2). Mash first assigns mixture proportions for each SNP onto each provided covariance matrix using maximum likelihood. Then, in a second step, each SNP's effect estimate undergoes shrinkage such that it is more aligned with the covariance structure inferred in the first step. The hypothesis-based covariance matrices are an important advantage mash offers for studying patterns of GxE: they allow hypothesis testing of specific environmental drivers for each SNP effect, and loadings on these matrices provide information on genome-wide patterns of SNP-environment interaction. Say that a SNP in FLC controls flowering in a temperature-dependent manner. In that case, that SNP effect could have a high mixture proportion, or mass, on a covariance matrix created using a temperature-based environmental cue, such as average temperature on the day prior to flowering. In our data, we would infer that the effect of that SNP on flowering was caused by a response to that or a correlated environmental cue.

The phenotypic correlations for green up date had moderate negative correlations between the Texas and North gardens, particularly in the Gulf and in Both subpopulations (Fig 1B). If these phenotypic correlations have a genetic component, then they might be controlled by SNP effects that differed in sign across these regions, SNP effects that are large in one garden or region and non-significant in others, or a combination of these patterns. Five hypothesis-based covariance matrices were selected by the greedy mash algorithm, one to two per subpopulation. Of these five, three had mass on them in mash models of the strong effects (Fig 2A, 2B). Two of these three matrices had negative covariances between sites between Texas and North gardens (Fig 2A), while one had all positive or near-zero covariances. SNP-associated phenotypic effects covaried with different weather-based cues in the Gulf & in Both subpopulations (Fig 2B). In total, 65% of the posterior weight of strong SNP effects in the mash model of Gulf green-up fell on a covariance matrix of daylength 14 days prior to the date of green-up. The covariance matrix for this weather cue was very similar to the pattern of phenotypic correlation for green-up in the Gulf subpopulation (Fig 1B; Fig 2A). Mash models of Midwest and Both subpopulation green-up did not include this covariance matrix; the Midwest had no weight on any hypothesis-based matrix, while Both subpopulations had non-zero weights on two additional hypothesis-based matrices, average temperature one day prior to green-up, and the day length change in seconds in the day prior to green-up (Fig 2B). The average temperature covariance matrix had negative covariances between Texas and North gardens, though not as strong as the negative phenotypic correlations seen in Both subpopulations (Fig 1B; Fig 2A). Only the Gulf subpopulation and Both subpopulations had mass on any hypothesis-based covariance matrices (Fig 2C).

For flowering date, distinct weather-based cues captured SNP-associated effect patterns in the Gulf and Midwest subpopulations. 33% of SNP effects on flowering in the Gulf subpopulation covaried with cumulative rainfall in the seven days prior to flowering (Fig 2D). 22.6% of SNP effects on flowering in the Midwest subpopulation contrived with day length change in the two days prior to flowering (Fig 2D), which had negative covariances between Texas and North gardens. Neither covariance matrix was selected for in Both subpopulations, and no hypothesis-based covariance matrices had mass in Both subpopulations (Fig 2E). In five of the six mash models of strong effects, the hypothesis-based covariance matrices captured a minority of the posterior weights of the strong effects (Figure 2C,E); the majority of this mass was on various canonical covariance matrices. These matrices included simple heterozygosity, with intermediate, positive covariances between all gardens, and garden-specific effects present only at one garden.

@@

We sought additional experimental support for our mash intervals using an independent pseudo-F2 mapping population created from Gulf & Midwest individuals and grown at the same sites (Fig. 4A,B). We conducted quantitative trait loci (QTL) mapping of flowering as functions of four environmental cues that we also used as covariance matrices in mash, and identified eight QTL for flowering date, six QTL for flowering GDD, ten QTL for flowering day length, and eight QTL flowering day length change, all of which showed QTL by environment interactions (SI Appendix, Fig. S3). All QTL for flowering overlapped one or more homologs from rice or A. thaliana with functionally validated roles in flowering (SI Appendix, Dataset 7). All flowering QTL intervals contained at least one SNP significant in at least one mash run at a log10-transformed Bayes Factor \> 2, or in the 1% tail of significance, whichever was stricter (SI Appendix, Dataset 8). We also looked for enrichments of mash SNPs in the 1% tail of significance (the 'mash 1% tail') within each QTL interval. At the 5% level, five QTL had enrichments of SNPs in the mash 1% tail. Overall, there were eleven significant enrichments (p \< 0.05, hypergeometric test) of SNPs in the mash 1% tail in the QTL intervals. Our QTL intervals had more enrichments of SNPs in the mash 1% tail than were found for all but eleven of these sets of random genomic intervals (Fig. 4C, p = 0.011). Thus, we were able to experimentally support our mash intervals with a QTL mapping experiment using a separate mapping population.

@@

As the climate and the natural environment change, it is increasingly critical to understand how patterns of gene-environment and plant-environment interactions will change in response. To do this, we must understand the current patterns of trait covariation across environments, the genetic underpinnings of these patterns, and the cases where this covariation can be altered. Here, we demonstrate that we can associate multiple patterns of GxE with specific genomic regions using a switchgrass diversity panel grown at eight common gardens, and also that we can assign specific SNP-associated patterns of GxE to both weather-based cues and to other, data-driven patterns. We use this approach to study GxE in both green-up and flowering phenological data in the deeply genetically diverged Gulf and Midwest populations of switchgrass.

Our analysis of green-up in the Gulf and in Both subpopulations revealed substantial antagonistic pleiotropy in effects between the Texas and North gardens (Figure 3A). This result supports theoretical models that local adaptation should involve antagonistic pleiotropy at the level of individual loci (22--25), and is the first experimental work using QTL mapping and GWAS across common gardens to find antagonistic pleiotropy to be common in small genomic regions (12, 36, 37).

Our analysis of flowering showed that the Gulf and Midwest subpopulations have two distinct photoperiod-related flowering responses: the Midwest subpopulation is day neutral, and flowering is cued primarily by a cumulative GDD threshold; in contrast, the Gulf subpopulation is photoperiod sensitive, and flowering is cued by the transition to shortening days. This result was supported by observations that expressing flowering date as a function of the day length at flowering increased its heritability in the Gulf subpopulation, while expressing flowering date as a function of cumulative GDD between green-up and flowering increased the heritability of flowering in the Midwest subpopulation (Fig. 1D). The genomic regions affecting flowering found in mash were also supported by QTL from an independent mapping population (Fig. 4C).

@@

\end{table}
```
![Figure 1. Characterization of green-up and flowering dates from the switchgrass diversity panel. (a) Map and trait histograms of green-up and flowering dates across two genetically distinct switchgrass subpopulations and eight common gardens. Purple represents individuals from the Midwest genetic subpopulation, and pink individuals from the Gulf subpopulation. Vertical dashed lines indicate the summer solstice. Common gardens are arranged in latitudinal order. (b) Phenotypic correlations between clonal replicates planted at eight common gardens, within and between two genetic subpopulations. (c) Narrow sense heritability of green-up and flowering within single common gardens (purple) and across all eight common gardens (green), within and between two genetic subpopulations. (d) Flow diagram of the methods applied to the green-up and flowering dates to jointly estimate SNP effects across all sites. Mash was fit to SNP effect data and used to find covariance matrices that improved the mash model likelihood using a large set of randomly selected, relatively unlinked SNP effects; this model was applied to a "strong" set of SNP effects with large effect sizes in the univariate GWAS.](images/Fig1_200dpi.png){#fig-map}

![Figure 2.](images/Figure_2_Covariance_matrices_and_posterior_weights.png){#fig-covar}

>>>>

title: Mapping Mixtures of Genotype-by-Weather Interactions in Switchgrass (Panicum virgatum)

@@

  The timing of vegetative and reproductive growth in plants ("phenological timings") depends on both genetic variation and environmental cues; thus, phenological traits can have important genotype-by-environment (GxE) interactions. Identifying the mixture and prevalence of GxE primed by weather prior to the phenological event (GxWeather) should aid prediction and manipulation of phenological timings. Here, we map GxWeather effects on phenological timings in two highly divergent switchgrass (*Panicum virgatum*) populations using repeated plantings of these populations at eight sites spanning the central United States. We distinguish GxWeather patterns covarying with interpretable weather-based cues from agnostic, site-based patterns. Most GxWeather effects belong to the latter category and do not covary with weather-based cues. However, 65% of effects on the timing of vegetative growth in the Gulf population covary with daylength 14 days prior to green-up date. 33% of effects on the timing of flowering in the Gulf population covary with cumulative rainfall in the seven days prior to flowering, while 22% of effects on flowering timing in the Midwest population covary with day length change two days prior to flowering. An independent pseudo-F2 cross of Gulf and Midwest individuals at the same sites mapped 23 additive QTLs for flowering, all with significant associations and eleven with enrichment of highly significant associations in our model of GxWeather effects. We demonstrate that we can identify QTL with GxWeather and assign them to specific weather-based cues or other patterns. Breeding for particular alleles at these loci could change flowering responsiveness to photoperiod and rainfall cues in switchgrass. More broadly, this approach could be used to identify genetic marker-environment interactions in any species with related populations phenotyped in multiple environments.

@@

  The timing of plant seasonal development (phenology) has major impacts on fitness because of the steep fitness cost of plant-environment mismatches. We infer the mixture of ways genetic effects on phenological traits depend on plant environment, focusing on how effects covary with weather prior to the phenological event (GxWeather). Most effects do not have GxWeather, but a minority do. GxWeather is population-specific. The majority of effects on the timing of vegetative growth in the Gulf population have GxWeather: these effects differ in sign across environments and covary with a photoperiod cue two weeks prior. A minority of effects on flowering date in the Gulf and Midwest populations had GxWeather, covarying with a cumulative rainfall and a photoperiod cue, respectively.

@@

\firstpage[9]{2}

Plant phenological timings are major components of plant fitness affected by multiple external environmental cues (e.g. degree of winter chilling, day length, temperature, and water availability) that signal existing or upcoming growing conditions [@bauerle_photoperiodic_2012; @andres2012genetic; @korner2010phenology]. Genetic responses to environmental cues determine the speed, timing, and energy apportioned to vegetative and reproductive growth and shape both the individual's lifespan and its lifetime production of viable seed. Day length (or photoperiod) is one of the most predictable environmental cues, and genetic sensitivity to photoperiod protects plants from potentially fatal consequences of phenological responses to temperature cues at the "wrong" time of year. However, the usefulness of specific environmental cues depends on both features of the environment, such as cue predictability and relevance, and the species' adaptive strategies @botero2015evolutionary. Species with wide natural distributions can have multiple distinct environmentally-cued phenological responses: for example, populations of sunflower (Helianthus annuus) exhibit day-neutral, facultative short day, and facultative long-day flowering responses, which vary with their environments [@blackman2013interacting; @henry2014transitions]. Distinct genetic responses in different environments are known as genotype by environment interactions, or GxE.

Flowering time, or the transition from vegetative to reproductive growth, is a common subject of GxE research [@blackman2013interacting; @henry2014transitions; @aagren2017adaptive; @brachi2010linkage; @dittmar2014flowering; @li2018genomic; @romero2017study], a key output of selection driving adaptation to local environments [@andres2012genetic; @wadgymar2017identifying; @blumel2015flowering], and a selection target for crop improvement to adapt crops to local or future environments @jung2009flowering. Changing flowering responsiveness to photoperiod cues has allowed geographic range expansion and increased yields in a number of cereal species [@turner2005pseudo; @faure2012mutation; @hung2012zmcct; @zakhrabekova2012induced; @yang2013oself3] and other crops [@pin2012multifaceted; @weller2019parallel]. Recent statistical advances in studying phenological GxE have involved determining critical environmental indices before the phenological event occurs, such as photothermal time within a critical growth window @li2018genomic. However, most studies of flowering GxE focus on finding a single, best fitting form of genotype-environment covariance, despite the key expectation that different genetic subpopulations, and even different genomic regions, have likely evolved distinct patterns of GxE. Additionally, despite theoretical predictions that local adaptation should involve antagonistic pleiotropy, or sign-changing GxE, at the level of individual loci [@levene1953genetic; @felsenstein1976theoretical; @kawecki2004conceptual; @hedrick1986genetic], previous work has found limited evidence of antagonistic pleiotropy [@des2013genotype; @wadgymar2017identifying]. However, this work has been limited by a known statistical bias that reduced detection of genetic effects that differ in sign [@des2013genotype; @anderson2013genetic; @anderson2011evolutionary]. Thus, despite substantial interest in the frequencies of various forms of GxE, the prevalence of antagonistic pleiotropy relative to other forms of GxE remains unknown.

Switchgrass (*Panicum virgatum*) is considered a short-day plant with reproductive development strongly linked to day of the year @mitchell1997predicting. However, as part of its wide environmental adaptation across the eastern half of North America, its photoperiodicity has been predicted to differ by plant latitude of origin [@parrish2005biology; @casler2004latitudinal]. We previously found divergent Midwest and Gulf genetic subpopulations of switchgrass with distinct sets of environmental adaptations, in that both populations had distinct genetic variation associated with each of two fitness proxies, biomass and survival @lovell2021genomic. The Midwest genetic subpopulation is primarily composed of individuals from the well-studied upland switchgrass ecotype [@porter1966analysis; @milano2016genetic], while the Gulf subpopulation has individuals from the well-studied lowland ecotype and the phenotypically intermediate coastal ecotype @lovell2021genomic.

Here, we estimate the ways genetic effects on phenological timings depend on weather prior to the phenological event. To do this, we phenotype a diversity panel of hundreds of switchgrass genotypes from the Midwest and Gulf subpopulations for the timing of vegetative and reproductive development at eight common garden locations spanning 17 degrees of latitude. These gardens cover the majority of the latitudinal and climatic range of switchgrass and capture the most comprehensive picture to date of the environmental variation this species encounters. We define a number of ways phenological timings might covary with weather (\@tbl-covar) and additional ways phenological timings might vary by site (\@SI-info), then jointly re-estimate genetic effects on these timings at all eight sites using the set of these covariance types that significantly improved the modeled log-likelihood when included (\@SI-info) @urbut2019flexible. The method we use for effect re-estimation, mash, allows us to identify and specify multiple ways genetic effects may covary with the environment, and does not have a statistical bias in detecting genetic effects with the same or opposite signs @10.1093/biostatistics/kxw041. To confirm our genetic mapping of GxWeather, we compare our re-estimated genetic effects to mapping results from an outbred pseudo-F2 cross grown at the same sites. Our analysis allow us to describe the weather cues and genetic variation affecting phenology in two divergent natural populations of switchgrass.

@@

Genotypes from the Gulf and Midwest subpopulations had distinct phenological timings and distinct patterns of phenological correlations across our eight common garden sites (@fig-map). At the three Texas common gardens (hereafter 'Texas' gardens), located within the natural range of the Gulf subpopulation, Gulf vegetative growth occurred before Midwestern vegetative growth, and Gulf flowering occurred after Midwestern flowering (@fig-map A). At the four northernmost common gardens (hereafter 'North' gardens), located within the natural range of the Midwest subpopulation, both Gulf vegetative growth and flowering occurred after Midwest vegetative growth and flowering. At the Oklahoma common garden, located near the natural range limits of both the Gulf and the Midwest subpopulations, Gulf and Midwest vegetative growth occurred over the same time period, and Gulf flowering occurred after Midwestern flowering (@fig-map A). These patterns led to strong negative phenotypic correlations for the timing of vegetative growth between the North and Texas gardens, particularly in the Gulf and across all individuals (hereafter, 'Both' subpopulations), and contributed to positive phenotypic correlations for the timing of flowering that had larger magnitudes at more northern gardens (@fig-map B).

@@

### GxWeather mixtures for vegetative and reproductive timing

We next looked for evidence of GxWeather by jointly re-estimating genetic effects across all eight common gardens using a model that allows effects to covary between gardens in many ways (@fig-map D, SI Appendix, Datasets 1-6). To do this, we used genome-wide association (GWAS) on site-specific BLUPs for the timing of vegetative growth and flowering to determine preliminary single nucleotide polymorphism (SNP) effect estimates at each garden. We then jointly modeled genetic effect estimates, and the mixture of ways these effects might covary, across all eight common gardens, using a two-part procedure. We first modeled genetic effects with different combinations of covariance types, using a subset of effects selected at random from all SNPs used in the GWAS. In these models, we specified many ways genetic effects could covary between gardens (as expanded on below) and used a greedy algorithm to iteratively select covariance structures that significantly improved the log likelihood of a model when included. Given the high correlation between some of the covariance patterns we tested, this approach allowed us to select only the subset of patterns of covariation that most improved the fit of the joint effect patterns present in the data. We then re-estimated the set of genetic effects that had the largest effect estimates in univariate GWAS (hereafter "strong" effects) using the set of significant covariance matrices that most improved the model fit on the random effects.

We created three categories of ways effects could covary across gardens: "canonical" covariance, with simple patterns of effects; "data-driven" covariances derived from common patterns of SNP effects in the data, and "GxWeather" covariances that captured the covariance of weather variables at specific times before the phenological event (Table 1, SI Appendix, Section S2). For example, GxWeather covariance between gardens in MI and MO, for rainfall in the seven days prior to flowering, captures the covariance in the amount of rainfall genotype A, B, and C receive in the seven days prior to their three flowering dates in MI and their three flowering dates in MO. If these six rainfall values are similar, covariance will be near one; if they are uncorrelated, covariance will be near zero; if there is sufficient rank changing in rainfall amounts between gardens for these genotypes, this covariance will be near negative one.

The GxWeather covariances allow hypothesis testing of specific weather variables as cues for each genetic effect. Say that a SNP in FLC controls flowering in a photoperiod-dependent manner. In that case, that SNP effect could have a high mixture proportion, or mass, on a covariance matrix created using a photoperiod-based environmental cue, such as day length two days prior to flowering. In our data, we would infer that the effect of that SNP on flowering was caused by a response to that or a correlated environmental cue. Overall, loadings of genetic effects on these GxWeather matrices in the model with re-estimated effects provide information on genome-wide patterns of SNP-environment interaction.

The phenotypic correlations for the timing of vegetative growth had moderate negative correlations between the Texas and North gardens, particularly in the Gulf and in Both subpopulations (@fig-map B). If these phenotypic correlations have a genetic component, then they might be controlled by SNP effects that differed in sign across these regions (effects with antagonistic pleiotropy), SNP effects that are large in one garden or region and non-significant in others (effects with amplification), or a combination of these patterns. Five GxWeather covariance matrices were selected by the greedy mash algorithm, one to two per subpopulation. Of these five, three had mass on them in mash models of the strong effects (@fig-covar A-B). Two of these three matrices had negative covariances between sites between Texas and North gardens (@fig-covar A), while one had all positive or near-zero covariances. SNP-associated phenotypic effects covaried with different weather-based cues in the Gulf & in Both subpopulations (@fig-covar B). In total, 65% of the posterior weight of strong SNP effects in the mash model of Gulf vegetative growth fell on a covariance matrix of daylength 14 days prior to the date of vegetative growth. The covariance matrix for this weather cue was very similar to the pattern of phenotypic correlation for the timing of vegetative growth in the Gulf subpopulation (@fig-map B; @fig-covar A). Mash models of the timing of vegetative growth in Midwest and Both subpopulations did not include this GxWeather covariance; the Midwest had no weight on any GxWeather covariance, while Both subpopulations had non-zero weights on two additional GxWeather covariance types, average temperature one day prior to green-up, and the day length change in seconds in the day prior to green-up (@fig-covar B). The average temperature covariance matrix had negative covariances between Texas and North gardens, though not as strong as the negative phenotypic correlations seen in Both subpopulations (@fig-map B; @fig-covar A). Only the Gulf subpopulation and Both subpopulations had mass on any GxWeather covariance matrices (@fig-covar C).

For flowering date, distinct GxWeather covariances captured SNP-associated effect patterns in the Gulf and Midwest subpopulations. 33% of SNP effects on flowering in the Gulf subpopulation covaried with cumulative rainfall in the seven days prior to flowering (Fig 2D). 22.6% of SNP effects on flowering in the Midwest subpopulation covaried with day length change in the two days prior to flowering (Fig 2D), which had negative covariances between Texas and North gardens. Neither covariance matrix was selected for in Both subpopulations, and no GxWeather covariance matrices had mass in Both subpopulations (Fig 2E). In five of the six mash models of strong effects, the GxWeather covariance matrices captured a minority of the posterior weights of the strong effects (Figure 2C,E); the majority of this mass was on various canonical covariance matrices. These matrices included simple heterozygosity, with intermediate, positive covariances between all gardens, and garden-specific effects present only at one garden.

@@

### Confirmation of effects on phenology using an independent mapping population

We sought additional experimental support for our significant effect re-estimates using an independent pseudo-F2 mapping population created from Gulf & Midwest individuals and grown at the same sites (Fig. 4A,B). We conducted quantitative trait loci (QTL) mapping of flowering as functions of four environmental cues that we also used as covariance matrices in mash, and identified eight QTL for flowering date, six QTL for flowering GDD, ten QTL for flowering day length, and eight QTL flowering day length change, all of which showed QTL by environment interactions (SI Appendix, Fig. S3). All QTL for flowering overlapped one or more homologs from rice or *A. thaliana* with functionally validated roles in flowering (SI Appendix, Dataset 7). All flowering QTL intervals contained at least one SNP significant in at least one mash run at a log10-transformed Bayes Factor \> 2, or in the 1% tail of significance, whichever was stricter (SI Appendix, Dataset 8). We also looked for enrichments of mash SNPs in the 1% tail of significance (the 'mash 1% tail') within each QTL interval. At the 5% level, five QTL had enrichments of SNPs in the mash 1% tail. Overall, there were eleven significant enrichments (p \< 0.05, hypergeometric test) of SNPs in the mash 1% tail in the QTL intervals. Our QTL intervals had more enrichments of SNPs in the mash 1% tail than were found for all but eleven of these sets of random genomic intervals (Fig. 4C, p = 0.011). Thus, we were able to experimentally support our significant effect re-estimates with a QTL mapping experiment using a separate mapping population.

@@

As the climate and the natural environment change, it is increasingly critical to understand how patterns of gene-environment and plant-environment interactions will change in response. To do this, we must understand the current patterns of trait covariation across environments, the genetic underpinnings of these patterns, and the cases where this covariation can be altered. Here, we demonstrate that we can associate multiple patterns of GxWeather with specific genomic regions using a switchgrass diversity panel grown at eight common gardens. We can assign genetic effects to both GxWeather patterns with interpretable weather-based cues, and to agnostic, site-based patterns. We use this approach to study GxWeather for the timings of vegetative and reproductive development in the deeply genetically diverged Gulf and Midwest subpopulations of switchgrass.

Our analysis of the timing of vegetative growth in the Gulf and in Both subpopulations revealed substantial antagonistic pleiotropy in effects between the Texas and North gardens (Figure 3A). This result supports theoretical models that local adaptation should involve antagonistic pleiotropy at the level of individual loci [@kawecki2004conceptual; @felsenstein1976theoretical; @levene1953genetic; @hedrick1986genetic], and is the first experimental work using QTL mapping and GWAS across common gardens to find antagonistic pleiotropy to be common in small genomic regions [@wadgymar2017identifying; @savolainen2013ecological; @lowry2019qtl].

Our analysis of the timing of flowering showed that the Gulf and Midwest subpopulations have distinct GxWeather: flowering timing in the Midwest subpopulation has photoperiod-related variation, in that flowering timing covaries with a day length change signal in the two days before flowering occurs. In contrast, the Gulf subpopulation does not have variation in flowering based on photoperiod. Instead, the Gulf subpopulation has variation in flowering that covaries with the rainfall that occurs in the week prior to flowering. The genomic regions affecting flowering that we re-estimated across all eight sites were also supported by QTL from an independent mapping population at these sites (Fig. 4C).

@@

\end{table}
```
![Characterization of green-up and flowering dates from the switchgrass diversity panel. (a) Map and trait histograms of green-up and flowering dates across two genetically distinct switchgrass subpopulations and eight common gardens. Purple represents individuals from the Midwest genetic subpopulation, and pink individuals from the Gulf subpopulation. Vertical dashed lines indicate the summer solstice. Common gardens are arranged in latitudinal order. (b) Phenotypic correlations between clonal replicates planted at eight common gardens, within and between two genetic subpopulations. (c) Narrow sense heritability of green-up and flowering within single common gardens (purple) and across all eight common gardens (green), within and between two genetic subpopulations. (d) Flow diagram of the methods applied to the green-up and flowering dates to jointly estimate SNP effects across all sites. Mash was fit to SNP effect data and used to find covariance matrices that improved the mash model likelihood using a large set of randomly selected, relatively unlinked SNP effects; this model was applied to a "strong" set of SNP effects with large effect sizes in the univariate GWAS.](images/Fig1_200dpi.png){#fig-map}

![Caption for this figure...](images/Figure_2_Covariance_matrices_and_posterior_weights.png){#fig-covar}
